---
title: "1st Assignment of Reproducible Research Course"
author: "Nicolas Corneau-Tremblay"
date: "7/18/2017"
output: github_document
---

## Preamble

```{r echo = FALSE}
setwd("/Users/nicot/Documents/Coursera/Data science/Reproducible/Ass1")
```

This document is my submission for the first assignment of Coursera - Reproducible Research course. To complete this asignment, I use the following libraries :

```{r message = FALSE}
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)
```


## 1 : Code for reading in the dataset and/or processing the data

The data for this assignment is from a personal activity monitoring device and is available on Coursera website.

```{r message = FALSE, cached = TRUE}
activity <- read_csv("activity.csv")
head(activity)
```

## 2 : Histogram of the total number of steps taken each day

Here is the histogram representing the total number of steps taken each day calculated on the entire sample. First, I create a new dataframe containing the total, the mean and the median number of steps taken for each day of the week. I then use this new dataframe to plot a histogram.

```{r message = FALSE, cached = TRUE}
day <- activity %>%
    group_by(date) %>%
    summarize(n.steps = sum(steps, na.rm = TRUE),
              mn.steps = mean(steps, na.rm = TRUE),
              md.steps = median(steps, na.rm = TRUE))


ggplot(day) +
    geom_histogram(aes(n.steps), fill = "orange") +
    xlab("Total number of steps taken") +
    ylab("Frequency")

```

## 3 : Mean and median number of steps taken each day

The **mean number of steps taken per day** is `r day %>% summarise(format(mean(n.steps, na.rm = TRUE), digits = 6, big.mark = ","))` and the **median number of steps taken per day** is `r day %>% summarise(format(median(n.steps, na.rm = TRUE), big.mark = ","))`.

## 4 : Time series plot of the average number of steps taken
```{r warning = FALSE}

ggplot(day) +
    geom_line(aes(date, mn.steps))
```

## 5 : The 5-minute interval that, on average, contains the maximum number of steps

```{r message = FALSE}
inter <- activity %>%
    group_by(interval) %>%
    summarize(mn.steps = mean(steps, na.rm = TRUE)) %>%
    arrange(desc(mn.steps))
```
The interval that contains, on average, the maximum number of steps is the `r head(inter, 1)[1]`th one of the day.

## 6 : Code to describe and show a strategy for imputing missing data

For the sake of this project, when the number of steps for an interval is missing, I impute the average number of steps taken in each interval of that day. If the entire day is missing, I then impute the whole sample mean for the interval. There is currently `r activity %>% subset(is.na(steps)) %>% summarise(format(n(), big.mark = ","))` missing values in the variable *steps*. The following code imputes the corresponding values :

```{r}
imputed <- activity %>%
    arrange(date, interval) %>%
    group_by(date) %>%
    mutate(day.mean = mean(steps, na.rm = TRUE)) %>%
    ungroup() %>%
    mutate(all.mean = mean(steps, na.rm = TRUE))

imputed <- imputed %>%
    mutate(imp.val = as.integer(round(coalesce(day.mean, all.mean)))) %>%
    mutate(steps = coalesce(steps, imp.val))

```

After imputation, there is `r imputed %>% subset(is.na(steps)) %>% summarise(format(n(), big.mark = ","))` missing value left in the new data set.

## 7 : Histogram of the total number of steps taken each day after missing values are imputed

```{r message = FALSE, cached = TRUE}
day.imp <- imputed %>%
    group_by(date) %>%
    summarize(n.steps = sum(steps, na.rm = TRUE),
              mn.steps = mean(steps, na.rm = TRUE),
              md.steps = median(steps, na.rm = TRUE))


ggplot(day.imp) +
    geom_histogram(aes(n.steps), fill = "orange") +
    xlab("Total number of steps taken (with imputation)") +
    ylab("Frequency")

```

## 8 : Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r message = FALSE, cached = TRUE}
week.day <- imputed %>%
    group_by(day = wday(date, label = TRUE, abbr = TRUE), interval) %>%
    summarize(mn.steps = mean(steps, na.rm = TRUE))

ggplot(week.day) +
    geom_smooth(aes(interval, mn.steps, group = day, color = day), se = FALSE) +
    labs(x = "5-minute interval", 
         y = "Average number of steps",
         color = "Days of the week")

```

## 9 : All of the R code needed to reproduce the results (numbers, plots, etc.) in the report

The R code is available throughout the document.






